{memer}: an R interface to the MEME Suite

### 2020-06-26 manuscript.Rmd
results draft to dan

---
title: "{memer}: an R interface to the MEME Suite"
author: 
 - name: Spencer L. Nystrom
   ORCID: 
 - name: Daniel J. McKay
output: github_document
---

```{r setup}
library(dremeR)
```

@@

When running a MEME Suite tool using `memer`, inputs are allowed in multiple
formats providing the user additional flexibility when performing motif analysis
(see "Core Utilities" for details). Of note, most outputs of `memer` tools can
be used as inputs to other `memer` tools, allowing easy construction of
"pipelines" (for example, discovering *de-novo* motifs with DREME, then finding
the genomic positions of those motifs in target regions using FIMO).
 - unlike current MEME tools (ex tomtom -> FIMO is hard to do)

@@

### Detecting a local MEME Suite Install (remove this section?)

`memer` requires the user to install the MEME Suite to the analysis environment
in which `memer` will be used. The `check_meme_install()` function allows users
to determine whether their R session is able to detect a local install of the
MEME Suite.
  - For functions that use MEME commandline tools, they need the location of 'meme/bin/' on disk, we provide utilities for making this detection simple to reduce friction of using the tool.
    - `check_meme_install` does user-friendly check that meme can be detected from R environment
    - Environment variable `MEME_PATH` (set & forget), `options(meme_path)` (configure for each script), `meme_path` argument, for each function call

@@

**Describe each tool briefly**

@@

When using list input to `memer` tools, names of list object members can be
passed to the `control` argument. This results in running the tool in
discriminative mode using those sequences as the control sequences for all other
list members. By passing multiple names to `control`, the sequences are
combined together to create a merged background set.

```{r}
sequences_list <- list("Static" = "seq",
                       "Increasing" = "seq",
                       "Decreasing" = "seq")

@@

## Metadata manipulation
 - `update_motifs()`
 - `force_best_match()`

## Data Visualization
 - tomtom best match
 - ame heatmaps by group

@@

**Somewhere say "memer can be used as a simple wrapper, but here we highlight its utility in chip-seq data analysis".**

>>>>

---
title: '{memer}: an R interface to the MEME Suite'
author:
- ORCID: null
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: word_document
---

```{r setup}
library(dremeR)
library(magrittr)
```

@@

When running a MEME Suite tool using `memer`, inputs are allowed in multiple
formats providing the user additional flexibility when performing motif analysis
(see "Core Utilities" for details). Of note, most outputs of `memer` tools can
be used as inputs to other `memer` tools, allowing easy construction of
pipelines (for example, discovering *de-novo* motifs with DREME, then finding
the genomic positions of those motifs in target regions using FIMO).
 - unlike current MEME tools (ex tomtom -> FIMO is hard to do)

@@

@@

**Describe each tool briefly?**
 - This could also be relegated to a table

@@

When using list input to `memer` tools, names of list object members can be
passed to the `control` argument. This results in running the tool in
discriminative mode using those sequences as the control sequences for all other
list members. By passing multiple names to `control`, the sequences are combined
together to create a merged background set. In this way, performing data-driven
discriminative analysis is just as simple as one using shuffled sequences as
background.

```{r, eval=F}
############
# Replace this for a real set of peaks 
sequences_list <- list("Static" = "seq",
                       "Increasing" = "seq",
                       "Decreasing" = "seq")
############

@@

## Metadata manipulation
`memer` is built on top of the `universalmotif` framework, which provides
methods for importing motifs from various file formats into a unified R data
representation complete with metadata information \cite{universalmotif}.
However, a major limitation of the `universalmotif` framework is that 
manipulating or filtering multiple `universalmotif` entries requires potentially
unfamiliar overhead for less experienced users, acting as a barrier to
performing potentially useful operations.

The `as_universalmotif_dataframe()` function from `memer` converts
`universalmotif` objects into a more familiar `data.frame` format which can be
manipulated using `tidyverse` or base R tools. These data.frames also contain a
special `motif` column which stores the `universalmotif` object of each entry.
Users can update columns in the data.frame which will be reflected in the
corresponding metadata entries of the `motif` by running `update_motifs()` after
manipulating the columns. These data.frames can be converted back into
`universalmotif` format using `as_universalmotif`, which also updates the
metadata using column values. The results from running the core MEME Suite
functions are stored in this format whenever appropriate, creating a consistent
framework for users to explore and manipulate motifs and their associated data values.

**This section is maybe too listy and could be fleshed out more / integrated above better**
Examples of common manipulations to motif databases which are aided by this
format include bulk renaming of entries, identification and removal of redundant
entries, and exploratory analysis of motif databases.

```{r}
# Example of renaming motifs w/ universalmotif syntax
library(universalmotif)
m1 <- create_motif()
m2 <- create_motif()
my_motifs <- list(m1,m2)

my_names <- c("motif one", "motif two")

for (i in 1:length(my_motifs)) {
  my_motifs[[i]]@name <- my_names[i]
}

# Example of renaming motifs w/ memer syntax
my_motifs <- list(m1,m2)
my_motifs %<>% 
  as_universalmotif_dataframe() %>% 
  dplyr::mutate(name = my_names) %>% 
  as_universalmotif()
```

This data format also allows users to integrate external data sources into their
analyses, for example, to filter a motif database to include only motifs for
genes which are expressed in their tissue of interest. Using motif databases
filtered in this way has been demonstrated in multiple contexts to provide
higher quality results when searching for novel regulators based on motif
content \cite{hector franco, eric olson, john stamm?, shep & greenleaf, klu worm
paper}.

## Data Visualization (section unfinished)

The MEME Suite provides a limited set of data visualizations which have limited
customizability. `memer` leverages the advantages of the R graphics environment
to provide novel data visualizations which are infinitely customizable by the user.

  - `view_tomtom_hits()` allows examination of tomtom results, allowing the user
  to evaluate the quality of the assigned match.

```{r}
# Example comparison of view_tomtom_hits results
plot_topHits <- example_tomtom %>% 
  dplyr::mutate(name = "Example Motif (Has Matches)") %>% 
  update_motifs() %>% 
  view_tomtom_hits(2)

plot_noMatch <- create_motif() %>% 
  as_universalmotif_dataframe() %>% 
  dplyr::mutate(name = "Example Motif (No Matches)") %>% 
  update_motifs() %>% 
  {
    x <- .
    x$tomtom <- list(NULL)
    return(x)
  } %>% 
  view_tomtom_hits()

cowplot::plot_grid(plotlist = list(plot_topHits[[1]], plot_noMatch[[1]]), ncol = 2, labels = "AUTO")
```

  - `ame_plot_heatmap()` allows comparison of AME results between multiple
  groups. Heirarchical ordering of hits shows commonalities and differences in enrichment between groups.
```{r}
# AME Heatmap example plots
example_ame$Decreasing %>% 
  ame_plot_heatmap()

example_ame %>% 
  dplyr::bind_rows(.id = "id") %>% 
  ame_plot_heatmap(group = id)
```

# Conclusion
 - `memer` is cool
 - use `memer`
 - it's on github (provide link)
 - manual is here:

# Notes

@@
